Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across Ontario’s Far North

### 2023-03-02 analysis/paper/paper.qmd
updates

---
title: "Ring of Fire Camera Detections Report"
author:
  - Valentin Lucet:
      # correspondence: "yes"
      email: valentin.lucet@gmail.com
      orcid: "0000-0003-0268-818X"
      institute: wlu
  - Frances Stewart:
      institute: wlu
      orcid: "0000-0001-9344-8346"
institute:
  - wlu:
      name: Wilfrid Laurier University
      # address: 23 Science Street, Eureka, Mississippi, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template_edited.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/landscape-ecology.csl" # Insert path for the bib-style
abstract: |
  Abstract TBW
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

<!-- The actual document text starts here: -->

@@

@@

- What this report is about
- Recycle the other introductions
- Importance of monitoring
- Data collection challenge
- New methods to the rescue
- Data challenges remain
- AI plan

@@

- Recycle the info from proposal

```{r}
#| label: spatial
#| echo: false

@@

# can <- rnaturalearth::ne_countries(country = "Canada", returnclass = "sf") %>% 
#   sf::st_crop(cams_sf)
# 
# can %>% 
#   ggplot2::ggplot() +
#   ggplot2::theme_bw() +
#   ggplot2::geom_sf() +
#   ggplot2::geom_sf(data=cams_sf)
```

@@

# Unknowns and bird breakdown
anns_wide_animals %>% dplyr::filter(species == "Unknown") %>% 
  dplyr::pull(id) %>% 
  stringr::str_sub(1,4) %>% table() %>% sort() 
anns_wide_animals %>% dplyr::filter(stringr::str_detect(species,"Bird")) %>% 
  dplyr::pull(id) %>% 
  stringr::str_sub(1,4) %>% table() %>% sort()

@@

We present a summary of the species identified and an ordination of presence absence data, using proportions of land cover classes observed in a radius of 1 km around each camera location. We used the 2015 Land Cover of Canada dataset.

@@

@@

events_plot_days <- anns_wide_animals_joined_mut %>% 
  # dplyr::mutate(ind = paste(species, sex, age, sep = "_")) %>% 
  # dplyr::select(camera_id, image_id, species, ymd) %>% 
  dplyr::select(plot_id, species, ymd) %>% 
  # dplyr::group_by(image_id, ind, ymd) %>% 
  # dplyr::summarise(ind_count=dplyr::n())
  dplyr::distinct()

@@

events_plot_tb <- events_plot %>% 
  # dplyr::group_by(camera_id) %>% 
  # dplyr::mutate(day_prop = day_count/sum(day_count)) # %>% 
  tidyr::pivot_wider(values_from = "day_count", names_from = "species") %>%
  dplyr::arrange(plot_id) %>%
  tibble::column_to_rownames("plot_id")

@@

# events <- events/rowSums(events)
events_plot_tb_filt_with_col <- events_plot_tb_filt %>% 
  tibble::rownames_to_column("plot_id")
```

@@

events_cam_days <- anns_wide_animals_joined_mut %>% 
  # dplyr::mutate(ind = paste(species, sex, age, sep = "_")) %>% 
  # dplyr::select(camera_id, image_id, species, ymd) %>% 
  dplyr::select(camera_id, species, month, ymd) %>% 
  # dplyr::group_by(image_id, ind, ymd) %>% 
  # dplyr::summarise(ind_count=dplyr::n())
  dplyr::distinct()

@@

events_tb_cam <- events_cam %>% 
  # dplyr::group_by(camera_id) %>% 
  # dplyr::mutate(day_prop = day_count/sum(day_count)) # %>% 
  tidyr::pivot_wider(values_from = "day_count", names_from = "species") %>%
  dplyr::arrange(camera_id) %>%
  tibble::column_to_rownames("camera_id")

@@

```{r}
#| label: ecoregions
#| echo: false

zones <- 
  sf::st_read("analysis/data/raw_data/ecozones/ECOREGN/LIO-2022-10-25/ECOREGION.shp",
              quiet=TRUE) %>% 
  sf::st_transform(crs = sf::st_crs(cams_sf))

# plot(zones$geometry)
```

```{r}
#| label: ecoregions-basic
#| echo: false

zones_filter <- zones %>%
  sf::st_filter(cams_sf)
# plot(zones_filter)
```

```{r}
#| label: ecoregions-ggplot-retrieved
#| echo: false

ggplot2::ggplot() +
  ggplot2::theme_bw() +
  ggplot2::geom_sf(data = zones_filter) +
  ggplot2::geom_sf(data = sf::st_jitter(cams_sf, 0.05), cex = 0.5, 
                   ggplot2::aes(col=is_retrieved)) +
  ggplot2::coord_sf(xlim = c(-89, -84),
                    ylim = c(50, 55)) +
  ggplot2::labs(col = "Camera \nRetrieved")
```

@@

regions_labels <- zones_filter %>% 
  sf::st_drop_geometry() %>% 
  dplyr::bind_cols(data.frame(x = c(-85, -85, -87.5),
                              y = c(54, 52, 52.7)))

base <- ggplot2::ggplot() +
  ggplot2::theme_bw()

for (pol in 1:3){
  base <- base + 
    ggplot2::geom_polygon(data = dplyr::filter(pols, L3 == pol),
                          ggplot2::aes(x=X, y=Y),
                          col = 1, fill = "grey90")
}

base +
  ggplot2::coord_equal(xlim = c(-88, -84.5),
                       ylim = c(50, 54)) +
  ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOREGION), 
                      data = regions_labels) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x=X, y=Y, group=plot_id),
                              cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x="Lat", y="Long")
```

@@

cams_sf_clust <- cams_sf %>% 
  # dplyr::filter(camera_id %in% names(clust_grp)) %>% 
  dplyr::left_join(grps_df, by = "camera_id") %>% 
  dplyr::filter(!(is.na(grp)))

ggplot2::ggplot(zones_filter) +
  ggplot2::theme_bw() +
  ggplot2::geom_sf() +
  ggplot2::geom_sf(data = sf::st_jitter(cams_sf_clust, 0.05), cex = 1, 
                   ggplot2::aes(col = grp)) +
  ggplot2::coord_sf(xlim = c(-89, -84),
                    ylim = c(50, 55))
```

@@

cams_sf_geo_clust <- cams_sf %>% 
  # dplyr::filter(camera_id %in% names(clust_grp)) %>% 
  dplyr::left_join(grps_geo_df, by = "camera_id") %>% 
  dplyr::filter(!(is.na(grp)))

@@

ggplot2::ggplot(zones_filter) +
  ggplot2::theme_bw() +
  ggplot2::geom_sf() +
  ggplot2::geom_sf(data = sf::st_jitter(cams_sf_geo_clust, 0.05), cex = 1, 
                   ggplot2::aes(col = grp)) +
  ggplot2::coord_sf(xlim = c(-89, -84),
                    ylim = c(50, 55))
```

@@

### Density heatmaps

@@

```{r}
cams_sf_sp %>%
  dplyr::filter(species == "Caribou", month == "Jun") %>%
  ggplot2::ggplot(ggplot2::aes(X, Y)) +
  ggplot2::geom_density_2d_filled(binwidth = 0.05) +
  ggplot2::coord_equal(xlim = c(-89, -84),
                       ylim = c(50, 55)) +
  ggplot2::theme(
    legend.position='none'
  )
```

```{r}
cams_sf_sp %>%
  dplyr::filter(species == "Caribou") %>%
  ggplot2::ggplot(ggplot2::aes(X, Y)) +
  ggplot2::geom_density_2d_filled(breaks = c(seq(0,1,by=0.1),5, 10, 20, 50,100),
                                  n=20) +
  ggplot2::facet_wrap(~month) +
  ggplot2::coord_equal(xlim = c(-89, -84),
                       ylim = c(50, 55)) +
  ggplot2::theme(
    legend.position='none'
  )
```

```{r}
# cams_sf_sp_sum %>%
#   # dplyr::filter(species == "Caribou") %>%
#   ggplot2::ggplot() +
#   ggplot2::geom_point(ggplot2::aes(x=X, y=Y, size=count, col =species)) +
#   ggplot2::facet_wrap(~month) +
#   ggplot2::coord_equal(xlim = c(-89, -84),
#                        ylim = c(50, 55))

@@

  gg <- ggplot2::ggplot(zones_filter) +
    ggplot2::theme_bw() +
    ggplot2::geom_sf() +
    ggplot2::geom_sf(data = dat,
                     ggplot2::aes(size = count)) +
    ggplot2::coord_sf(xlim = c(-89, -84),
                      ylim = c(50, 54)) +
    ggplot2::facet_wrap(~month) +
    ggplot2::labs(size="Days \ndetected")

  ggplot2::ggsave(gg, filename = paste0("analysis/figures/",sp,".png"))
}
```

```{r}
# cams_sf_sp_sum %>%
#   dplyr::filter(species == "Caribou") %>%
#   ggplot2::ggplot() +
#   ggplot2::geom_raster(ggplot2::aes(x=X, y=Y, z=count)) +
#   ggplot2::facet_wrap(~month) +
#   ggplot2::coord_equal(xlim = c(-89, -84),
#                        ylim = c(50, 55))
```

@@

- Thank the volunteers for help with tagging

<!-- The following line inserts a page break  -->

@@

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

@@

<!-- ### Colophon

@@

```{r}
# -------------------------------------------------------------------------

>>>>

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across Ontario’s Far North"
author:
  - Valentin Lucet:
      # correspondence: "yes"
      email: valentin.lucet@gmail.com
      orcid: "0000-0003-0268-818X"
      institute: wlu
  - Frances Stewart:
      institute: wlu
      orcid: "0000-0001-9344-8346"
institute:
  - wlu:
      name: Wilfrid Laurier University
      # address: 23 Science Street, Eureka, Mississippi, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template_modified.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/landscape-ecology.csl" # Insert path for the bib-style
abstract: |
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed in an heterogeneous sampling area straddling two of Northern Ontario’s ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We've recently developed an open-source workflow for camera trap image identification, access, and storage, which we are using to identify images, with next steps involving the development of an image classifier.

# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

<!-- The actual document text starts here:

@@

\newpage -->

@@

The James Bay Lowlands in Northern Ontario hold some of the world’s last intact wild places but it is under pressure from resource extraction and global warming. Our understanding of basic biodiversity measures like species occurrence, abundance, diversity, and density for the region are still rudimental but needed to weigh ecological-economical decision tradeoffs.   
Traditionally, quantifying and monitoring biodiversity in northern regions required expensive and time-consuming field surveys. But today, the use of non-intrusive methods, such as arrays of 100s of camera traps, has significantly changed field practices for reliably quantifying wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification, and processing, of image datasets.

Machine learning methods, such as deep neural networks, are now in use to automate the process of species identification and help in that endeavour. We aim to evaluate the viability of deep neural networks for camera trap surveys currently taking place in Northern Ontario’s heterogeneous landscapes, including the Hudson Bay Lowlands and Ontario Shield ecozones, as variables like weather, seasonality, vegetation type, and species characteristics can impact the performance of these methods. There are currently no available machine learning models for identifying species from images in northern regions and developing one would be beneficial as a monitoring and modeling tool for northern boreal ecosystems.

The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed in an heterogeneous sampling area straddling two of Northern Ontario’s ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We've recently developed an open-source workflow for camera trap image identification, access, and storage, which we are using to identify images, with next steps involving the development of an image classifier.

@@

The camera deployment protocol follows a spatially balanced hierarchical design that provides an even distribution of sample sites across biotic and abiotic conditions within the study area. The study sites span three ecoregions and two ecozones. The network was established by CWS in 2020 and seasonally equipped in a rolling design with 500 autonomous recording units (ARUs) to monitor audible species diversity paired with site-specific on-ground vegetation imagery. In 2022, 232 unbaited wildlife camera traps (CTs) were added to a subset of study sites to quantify mammalian occurrence and diversity. @fig-deployment shows the location of the and their current status. retrieved cameras were retrieved in September 2022.

```{r}
#| label: spatial-basics
#| echo: false

@@

zones <- 
  sf::st_read("analysis/data/raw_data/ecozones/ECOREGN/LIO-2022-10-25/ECOREGION.shp",
              quiet=TRUE, ) %>% 
  sf::st_transform(crs = sf::st_crs(cams_sf))

zones_filter <- zones %>%
  sf::st_filter(cams_sf)

regions_labels <- zones_filter %>% 
  sf::st_drop_geometry() %>% 
  dplyr::bind_cols(data.frame(x = c(-88.5, -84, -88.5),
                              y = c(55, 52, 52.6)))

map <- ggmap::get_stamenmap( bbox = c(left = -90, bottom = 50, right = -83, top = 56), 
                             zoom = 9, maptype = "terrain-background")
```

```{r}
#| label: fig-deployment
#| echo: false
#| fig-cap: "Camera deployment status"

base <- ggmap::ggmap(map) +
  ggplot2::labs(x = "Longitude", y = "Latitude") +
  # ggplot2::geom_polygon(data = pols, ggplot2::aes(x=X, y=Y), col = 1, fill = NA)
  ggplot2::geom_path(data = pols, ggplot2::aes(x=X, y=Y), col = 1, linewidth=0.2)

p <- base +
  ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOREGION),inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords, cex = 1, shape=21, col="black",
                      ggplot2::aes(x=X, y=Y, fill=is_retrieved), 
                      inherit.aes = FALSE) +
  ggplot2::labs(fill = "Camera \nRetrieved")
p
```

@@

# Unknowns and bird breakdown
# anns_wide_animals %>% dplyr::filter(species == "Unknown") %>% 
#   dplyr::pull(id) %>% 
#   stringr::str_sub(1,4) %>% table() %>% sort() 
# anns_wide_animals %>% dplyr::filter(stringr::str_detect(species,"Bird")) %>% 
#   dplyr::pull(id) %>% 
#   stringr::str_sub(1,4) %>% table() %>% sort()

@@

#### Event definition

- days

#### Ordination and clustering

- two types of ordination: pa and events
- two clustering: basics and geo

@@

- Most common species

@@

events_plot_days <- anns_wide_animals_joined_mut %>% 
  # dplyr::mutate(ind = paste(species, sex, age, sep = "_")) %>% 
  dplyr::select(plot_id, species, ymd) %>% 
  dplyr::distinct()

@@

events_plot_tb <- events_plot %>% 
  tidyr::pivot_wider(values_from = "day_count", names_from = "species") %>%
  dplyr::arrange(plot_id) %>%
  tibble::column_to_rownames("plot_id")

@@

events_plot_tb_filt_with_col <- events_plot_tb_filt %>% 
  tibble::rownames_to_column("plot_id")
```

@@

events_cam_days <- anns_wide_animals_joined_mut %>% 
  dplyr::select(camera_id, species, month, ymd) %>% 
  dplyr::distinct()

@@

events_tb_cam <- events_cam %>% 
  tidyr::pivot_wider(values_from = "day_count", names_from = "species") %>%
  dplyr::arrange(camera_id) %>%
  tibble::column_to_rownames("camera_id")

@@

@@

base2 <- ggplot2::ggplot() +
  ggplot2::theme_bw()
for (pol in 1:3){
  base2 <- base2 + 
    ggplot2::geom_polygon(data = dplyr::filter(pols, L3 == pol),
                          ggplot2::aes(x=X, y=Y),
                          col = 1, fill = "grey90")
}

base2 +
  ggplot2::coord_equal(xlim = c(-88.5, -84),
                       ylim = c(50, 54)) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x=X, y=Y, group=plot_id),
                              cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x="Lat", y="Long")
```

@@

cams_sf_clust <- cams_sf_coords %>% 
  # dplyr::filter(camera_id %in% names(clust_grp)) %>% 
  dplyr::left_join(grps_df, by = "camera_id") %>% 
  dplyr::filter(!(is.na(grp)))

base_cut <- base +
  ggplot2::coord_map(xlim = c(-89, -84),
                     ylim = c(50, 54.5))

base_cut +
  ggplot2::geom_point(data = cams_sf_clust, cex = 1, shape=21, col="black",
                      ggplot2::aes(x=X, y=Y, fill=grp), 
                      inherit.aes = FALSE) +
  ggplot2::labs(fill="Group")
```

@@

cams_sf_geo_clust <- cams_sf_coords %>% 
  # dplyr::filter(camera_id %in% names(clust_grp)) %>% 
  dplyr::left_join(grps_geo_df, by = "camera_id") %>% 
  dplyr::filter(!(is.na(grp)))

@@

base_cut +
  ggplot2::geom_point(data = cams_sf_geo_clust, cex = 1, shape=21, col="black",
                      ggplot2::aes(x=X, y=Y, fill=grp), 
                      inherit.aes = FALSE) +
  ggplot2::labs(fill="Group")
```

@@

### Density contourmaps

@@

```{r}
#| label: heatmap
#| echo: false

cams_sf_sp_filt <- cams_sf_sp %>% 
  dplyr::filter(species == "Caribou", month == "Jun")

base_cut +
  ggplot2::geom_density2d_filled(
    data = cams_sf_sp_filt, ggplot2::aes(X, Y),
    binwidth = 0.05, alpha = 0.5)
```

```{r}
#| label: heatmap-facets
#| echo: false

cams_sf_sp_caribou <- cams_sf_sp %>%
  dplyr::filter(species == "Caribou")

base_cut +
  ggplot2::geom_density2d_filled(
    data = cams_sf_sp_caribou, ggplot2::aes(X, Y),
    alpha = 0.5,breaks = c(seq(0.001,1,by=0.1),5, 10, 20, 50,100), n=20) +
  ggplot2::facet_wrap(~month) +
  ggplot2::coord_equal(xlim = c(-89, -84),
                       ylim = c(50, 55)) +
  ggplot2::theme(
    legend.position='none'
  )
```

```{r}
#| label: bubbleplots
#| echo: false

# cams_sf_sp_sum %>%
#   # dplyr::filter(species == "Caribou") %>%
#   ggplot2::ggplot() +
#   ggplot2::geom_point(ggplot2::aes(x=X, y=Y, size=count, col =species)) +
#   ggplot2::facet_wrap(~month) +
#   ggplot2::coord_equal(xlim = c(-89, -84),
#                        ylim = c(50, 55))

@@

  gg <- base_cut +
    ggplot2::geom_point(data = dat, col="black",
                        ggplot2::aes(x=X, y=Y, size = count), 
                        inherit.aes = FALSE) +
    ggplot2::facet_wrap(~month) +
    ggplot2::labs(size="Days \ndetected")

  ggplot2::ggsave(gg, filename = paste0("analysis/figures/",sp,"_bplot.png"))
}
```

![alt text](../figures/Caribou_bplot.png "Caribou")
```{r}
#| label: heatmaps
#| echo: false

  cams_sf_sp_spe <- cams_sf_sp %>%
    dplyr::filter(species == sp)

  gg <- base_cut +
    ggplot2::geom_density2d_filled(
      data = cams_sf_sp_spe, ggplot2::aes(X, Y),
      alpha = 0.5,breaks = c(seq(0.001,1,by=0.1),5, 10, 20, 50,100), n=20) +
    ggplot2::facet_wrap(~month) +
    ggplot2::coord_equal(xlim = c(-89, -84),
                         ylim = c(50, 55)) +
    ggplot2::theme(
      legend.position='none'
    )

  ggplot2::ggsave(gg, filename = paste0("analysis/figures/",sp,"_heatmap.png"))
}
```
![alt text](../figures/Caribou_heatmap.png "Caribou")

```{r}
#| label: plot-tests
#| echo: false

# cams_sf_sp_sum %>%
#   dplyr::filter(species == "Caribou") %>%
#   ggplot2::ggplot() +
#   ggplot2::geom_raster(ggplot2::aes(x=X, y=Y, z=count)) +
#   ggplot2::facet_wrap(~month) +
#   ggplot2::coord_equal(xlim = c(-89, -84),
#                        ylim = c(50, 55))
```

@@

- Thank the volunteers for help with tagging.

<!-- The following line inserts a page break  -->
<!--
\newpage

@@

The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  
-->

@@

 ### Colophon

@@

<!-- 
```{r}
# -------------------------------------------------------------------------
